---
title: "Comparing Census Geographies"
subtitle: <h4 style="font-style:normal">CRD 150 - Quantitative Methods in Community Research</h4>
date: <h4 style="font-style:normal">Winter 2020</h4>
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cosmo
    code_folding: show
---


<style>
p.comment {
background-color: #DBDBDB;
padding: 10px;
border: 1px solid black;
margin-left: 25px;
border-radius: 5px;
font-style: italic;
}

.figure {
   margin-top: 20px;
   margin-bottom: 20px;
}

h1.title {
  font-weight: bold;
  font-family: Arial;  
}

h2.title {
  font-family: Arial;  
}

</style>


<style type="text/css">
#TOC {
  font-size: 13px;
  font-family: Arial;
}
</style>

\

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Your final project involves studying an interesting question about a U.S. community.  We define community as a [city](https://www.census.gov/geo/reference/gtc/gtc_place.html), or [county](https://www.census.gov/geo/reference/gtc/gtc_cou.html).  What is the spatial extent of these communities and how do they compare in size, scale, and area?  Below is a map comparing the Sacramento Metropolitan Area (Grey), Sacramento City (Red), and Sacramento County (Blue).  You can hide and add  layers to see how the geographies compare.   

We will also work a lot with neighborhoods in this class.  This either means using Census [tracts](https://www.census.gov/geo/reference/gtc/gtc_ct.html) or [blocks](https://www.census.gov/geo/reference/gtc/gtc_block.html), and also sometimes zip codes (or [ZCTAs](https://www.census.gov/geo/reference/zctas.html)).  Below is a map comparing Davis zip codes (Grey), tracts (blue), and blocks (red). Can you find the tract you  currently live in? What about the block?

One thing you'll have to be careful about if you're interested in comparing tracts over time - [boundaries change](https://www.census.gov/geo/reference/gtc/gtc_boundarychange.html).  We show a map comparing Davis tracts in 2000 (red) and 2010 (black). You'll notice that some tracts in 2000 get split up.  Why?  Because there were enough significant population changes within the tract from 2000 to 2010 to warrant a split.  If you are curious, for example, why the tract in East Davis got split up into 3 tracts, check out this side-by-side [Social Explorer map](https://www.socialexplorer.com/65d3c9db2b/view) showing changes in population size in the 3 tracts from 2000 to 2010. I also embedded that map below. Can you tell what happened?

The code producing each map is shown after the map. 

<div style="margin-bottom:25px;">
</div>
## **Sacramento metro, county and city**
<div style="margin-bottom:100px;">
</div>

```{r include = FALSE, warning = FALSE, message = FALSE}
library(sf)
library(tidycensus)
census_api_key("b81d373d6e785ecbc489de1fc862aef424d0a63a")
library(tigris)
options(tigris_class = "sf")
library(tidyverse)
library(rmapshaper)

          
cb <- core_based_statistical_areas(cb = TRUE)
pl <- places(state = "CA", cb = TRUE)
cnty <- counties(state ="CA", cb = TRUE)
ca.tracts <- tracts(state ="CA", cb = TRUE)
yolo.blocks <-blocks(state="CA", county = "Yolo")
yolo.block.groups <-block_groups(state="CA", county = "Yolo", cb = TRUE)
zips <- zctas(cb=TRUE, state = "CA")
ca.tracts.2000 <- tracts(state ="CA", cb = TRUE, year = 2000)
ca.tracts.2000<-st_transform(ca.tracts.2000, crs=st_crs(ca.tracts))

sac.metro <- filter(cb, NAME =="Sacramento--Roseville--Arden-Arcade, CA")
sac.city <- filter(pl, NAME == "Sacramento")
sac.county <- filter(cnty, NAME == "Sacramento")
davis.city <- filter(pl, NAME == "Davis")

davis.tracts <- ms_clip(ca.tracts, davis.city, remove_slivers = TRUE) 
davis.blocks <- ms_clip(yolo.blocks, davis.city, remove_slivers = TRUE) 
davis.block.groups <- ms_clip(yolo.block.groups, davis.city, remove_slivers = TRUE) 
davis.zip <- ms_clip(zips, davis.city, remove_slivers = TRUE) 
davis.tracts.2000 <- ms_clip(ca.tracts.2000, davis.city, remove_slivers = TRUE) 

```


```{r leaflet, fig.align="center", echo = FALSE, warning = FALSE, message = FALSE}
library(leaflet)

sac.map <- leaflet(sac.metro) %>%
  # Base groups
  addTiles(urlTemplate = "http://{s}.tile.openstreetmap.org/{z}/{x}/{y}.png") %>%
  # Overlay groups
  addPolygons(color="grey", group = "Metro", weight = 1, smoothFactor = 0.5,
    opacity = 1.0, fillOpacity = 0.5) %>%
  addPolygons(data = sac.county, color = "blue", group = "County", weight = 1, smoothFactor = 0.5,
    opacity = 1.0, fillOpacity = 0.5) %>%
  addPolygons(data = sac.city, color = "red", group = "City", weight = 1, smoothFactor = 0.5,
    opacity = 1.0, fillOpacity = 0.5) %>%
  # Layers control
  addLayersControl(
    overlayGroups = c("Metro", "City", "County"),
    options = layersControlOptions(collapsed = FALSE)
  )
sac.map
```

<div style="margin-bottom:100px;">
</div>



```{r eval = FALSE, warning = FALSE, message = FALSE}
#Load in required packages
library(tigris)
options(tigris_class = "sf")
library(tidyverse)
library(rmapshaper)

#extract metro, counties and cities          
cb <- core_based_statistical_areas(cb = TRUE)
pl <- places(state = "CA", cb = TRUE)
cnty <- counties(state ="CA", cb = TRUE)

#subset to Sacramento metro, county, and city
sac.metro <- filter(cb, NAME =="Sacramento--Roseville--Arden-Arcade, CA")
sac.city <- filter(pl, NAME == "Sacramento")
sac.county <- filter(cnty, NAME == "Sacramento")
```

```{r eval = FALSE, warning = FALSE, message = FALSE}
library(leaflet)

sac.map <- leaflet(sac.metro) %>%
  # Base groups
  addTiles(group = "OSM (default)") %>%
  # Overlay groups
  addPolygons(color="grey", group = "Metro", weight = 1, smoothFactor = 0.5, 
              opacity = 1.0, fillOpacity = 0.5) %>%
  addPolygons(data = sac.city, color = "red", group = "City", weight = 1, 
              smoothFactor = 0.5, opacity = 1.0, fillOpacity = 0.5) %>%
  addPolygons(data = sac.county, color = "blue", group = "County", weight = 1, 
              smoothFactor = 0.5, opacity = 1.0, fillOpacity = 0.5) %>%
  # Layers control
  addLayersControl(
    overlayGroups = c("Metro", "City", "County"),
    options = layersControlOptions(collapsed = FALSE)
  )
sac.map
```

<div style="margin-bottom:25px;">
</div>
## **Davis ZCTAs, tracts, block groups, and blocks**
<div style="margin-bottom:100px;">
</div>



```{r echo = FALSE, warning = FALSE, message = FALSE}

davis.map <- leaflet(davis.zip) %>%
  # Base groups
  addTiles(urlTemplate = "http://{s}.tile.openstreetmap.org/{z}/{x}/{y}.png") %>%
  # Overlay groups
  addPolygons(color="black", group = "Zipcode", weight = 1, smoothFactor = 0.5,
    opacity = 1.0, fillOpacity = 0.5) %>%
  addPolygons(data = davis.tracts, color = "blue", group = "Tract", weight = 1, smoothFactor = 0.5,
    opacity = 1.0, fillOpacity = 0.5) %>%
  addPolygons(data = davis.block.groups, color = "green", group = "Block group", weight = 1, smoothFactor = 0.5, opacity = 1.0, fillOpacity = 0.5) %>%
  addPolygons(data = davis.blocks, color = "red", group = "Block", weight = 1, smoothFactor = 0.5,
    opacity = 1.0, fillOpacity = 0.5) %>%
  # Layers control
  addLayersControl(
    overlayGroups = c("Zipcode", "Tract", "Block group", "Block"),
    options = layersControlOptions(collapsed = FALSE)
  )
davis.map
```

<div style="margin-bottom:100px;">
</div>


```{r eval = FALSE, warning = FALSE, message = FALSE}
#Load required packages
library(tidycensus)
census_api_key("INSERT YOUR KEY HERE")
library(tigris)
options(tigris_class = "sf")
library(tidyverse)
library(rmapshaper)

#Get cities, tracts, blocks, and ZCTAs
pl <- places(state = "CA", cb = TRUE)
ca.tracts <- tracts(state ="CA", cb = TRUE)
yolo.block.groups <-block_groups(state="CA", county = "Yolo", cb = TRUE)
yolo.blocks <-blocks(state="CA", county = "Yolo")
zips <- zctas(cb=TRUE, state = "CA")

#Davis city
davis.city <- filter(pl, NAME == "Davis")

#Clip tracts, blocks and ZCTAs to Davis city boundary
davis.tracts <- ms_clip(ca.tracts, davis.city, remove_slivers = TRUE) 
davis.block.groups <- ms_clip(yolo.block.groups, davis.city, remove_slivers = TRUE) 
davis.blocks <- ms_clip(yolo.blocks, davis.city, remove_slivers = TRUE) 
davis.zip <- ms_clip(zips, davis.city, remove_slivers = TRUE) 
```

```{r eval = FALSE, warning = FALSE, message = FALSE}
library(leaflet)

davis.map <- leaflet(davis.zip) %>%
  # Base groups
  addTiles(urlTemplate = "http://{s}.tile.openstreetmap.org/{z}/{x}/{y}.png") %>%
  # Overlay groups
  addPolygons(color="black", group = "Zipcode", weight = 1, smoothFactor = 0.5, 
              opacity = 1.0, fillOpacity = 0.5) %>%
  addPolygons(data = davis.tracts, color = "blue", group = "Tract", weight = 1, 
              smoothFactor = 0.5, opacity = 1.0, fillOpacity = 0.5) %>%
  addPolygons(data = davis.block.groups, color = "green", group = "Block group", 
              weight = 1, smoothFactor = 0.5, opacity = 1.0, fillOpacity = 0.5) %>%
  addPolygons(data = davis.blocks, color = "red", group = "Block", weight = 1, 
              smoothFactor = 0.5, opacity = 1.0, fillOpacity = 0.5) %>%
  # Layers control
  addLayersControl(
    overlayGroups = c("Zipcode", "Tract", "Block group", "Block"),
    options = layersControlOptions(collapsed = FALSE)
  )
davis.map
```


<div style="margin-bottom:25px;">
</div>
## **Davis tracts in 2000 and 2010**
<div style="margin-bottom:100px;">
</div>



```{r echo = FALSE, warning = FALSE, message = FALSE}

davis.map2 <- leaflet(davis.tracts) %>%
  # Base groups
  addTiles(urlTemplate = "http://{s}.tile.openstreetmap.org/{z}/{x}/{y}.png") %>%
  # Overlay groups
  addPolygons(color="black", group = "2010", weight = 1, smoothFactor = 0.5,
    opacity = 1.0, fillOpacity = 0.5) %>%
  addPolygons(data = davis.tracts.2000, color = "red", group = "2000", weight = 1, smoothFactor = 0.5,
    opacity = 1.0, fillOpacity = 0.5) %>%
  # Layers control
  addLayersControl(
    overlayGroups = c("2010", "2000"),
    options = layersControlOptions(collapsed = FALSE)
  )
davis.map2
```

<div style="margin-bottom:100px;">
</div>

Why were there changes? The following map from Social Explorer shows total population sizes in Davis tracts in 2010 boundaries in 2000 and 2010.  Toggle between 2000 to 2010 by selecting Map 1 and Map 2.

<div style="margin-bottom:100px;">
</div>

<p align="center"><iframe frameborder="0" scrolling="no" marginheight="0" marginwidth="0" src="https://www.socialexplorer.com/65d3c9db2b/embed" width="800" height="600" allowfullscreen="true" webkitallowfullscreen="true" mozallowfullscreen="true"></iframe></iframe></p>

<div style="margin-bottom:100px;">
</div>

```{r eval = FALSE, warning = FALSE, message = FALSE}
#Load in required packages
library(tidycensus)
census_api_key("b81d373d6e785ecbc489de1fc862aef424d0a63a")
library(tigris)
options(tigris_class = "sf")
library(tidyverse)
library(rmapshaper)

#Bring in cities, tracts in 2000, and tracts in 2010
pl <- places(state = "CA", cb = TRUE)
ca.tracts.2010 <- tracts(state ="CA", cb = TRUE)
ca.tracts.2000 <- tracts(state ="CA", cb = TRUE, year = 2000)
# Need to reproject 2000 tracts to 2010 tract CRS
ca.tracts.2000<-st_transform(ca.tracts.2000, crs=st_crs(ca.tracts))

#Get City of Davis
davis.city <- filter(pl, NAME == "Davis")

#Clip tracts to Davis boundaries
davis.tracts.2010 <- ms_clip(ca.tracts.2010, davis.city, remove_slivers = TRUE) 
davis.tracts.2000 <- ms_clip(ca.tracts.2000, davis.city, remove_slivers = TRUE) 

```

```{r eval = FALSE, warning = FALSE, message = FALSE}
library(leaflet)

davis.map2 <- leaflet(davis.tracts.2010) %>%
  # Base groups
  addTiles(urlTemplate = "http://{s}.tile.openstreetmap.org/{z}/{x}/{y}.png") %>%
  # Overlay groups
  addPolygons(color="black", group = "2010", weight = 1, smoothFactor = 0.5,
    opacity = 1.0, fillOpacity = 0.5) %>%
  addPolygons(data = davis.tracts.2000, color = "red", group = "2000", weight = 1, 
              smoothFactor = 0.5, opacity = 1.0, fillOpacity = 0.5) %>%
  # Layers control
  addLayersControl(
    overlayGroups = c("2010", "2000"),
    options = layersControlOptions(collapsed = FALSE)
  )
davis.map2
```


***

Website created and maintained by [Noli Brazil](https://nbrazil.faculty.ucdavis.edu/)

