---
title: "Eligible Metros, Cities and Counties"
subtitle: <h4 style="font-style:normal">CRD 150 - Quantitative Methods in Community Research</h4>
date: <h4 style="font-style:normal">Fall 2018</h4>
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cosmo
---


<style>
p.comment {
background-color: #DBDBDB;
padding: 10px;
border: 1px solid black;
margin-left: 25px;
border-radius: 5px;
font-style: italic;
}

h1.title {
  font-weight: bold;
}

</style>

\


Your final project involves studying an interesting question about a U.S. community.  We define community as a [metropolitan area](https://www.census.gov/geo/reference/gtc/gtc_cbsa.html#mesa), [city](https://www.census.gov/geo/reference/gtc/gtc_place.html), or [county](https://www.census.gov/geo/reference/gtc/gtc_cou.html).  You must select a metropolitan area, city, or county with a population size greater than 250,000.  The eligible metro areas, cities, and counties using 2012-2016 American Community Survey estimates are shown below in order of population size.

<div style="margin-bottom:25px;">
</div>
#Metropolitan Areas

```{r echo=  FALSE, warning = FALSE, message = FALSE}
options(knitr.kable.NA = '')
library(knitr)
library(kableExtra)
library(sf)
library(tidyverse)
dt <- read_csv("/Users/noli/Documents/UCD/teaching/CRD150/project/nhgis0103_csv/metro_area.csv")
kable(dt) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "responsive")) %>%
    column_spec(2, width = "15em")
```

<div style="margin-bottom:25px;">
</div>
#Cities 
```{r echo=  FALSE, warning = FALSE, message = FALSE}
options(knitr.kable.NA = '')
library(knitr)
library(kableExtra)
library(sf)
library(tidyverse)
dt <- read_csv("/Users/noli/Documents/UCD/teaching/CRD150/project/nhgis0103_csv/cities.csv")
kable(dt) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "responsive")) %>%
    column_spec(4, width = "15em")
```

<div style="margin-bottom:25px;">
</div>
#Counties

```{r echo=  FALSE, warning = FALSE, message = FALSE}
options(knitr.kable.NA = '')
library(knitr)
library(kableExtra)
library(sf)
library(tidyverse)
dt <- read_csv("/Users/noli/Documents/UCD/teaching/CRD150/project/nhgis0103_csv/counties.csv")
kable(dt) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "responsive")) %>%
    column_spec(4, width = "15em")
```
